Analytic Doppler Imaging

### 2019-08-02 tex/gitlinks.tex
more words

\newcommand{\codelink}[1]{\href{https://github.com/user/repo/blob/7c3f626863f61d2a592a381e10d381862ebcb9fb/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/user/repo/blob/7c3f626863f61d2a592a381e10d381862ebcb9fb/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/user/repo/blob/7c3f626863f61d2a592a381e10d381862ebcb9fb/tex/proofs/#1.ipynb}{\raisebox{-0.1em}{\prooficon}}}

>>>>

\newcommand{\codelink}[1]{\href{https://github.com/user/repo/blob/93997d2ef089c7be911a4ebbb4e4d41963d4d260/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/user/repo/blob/93997d2ef089c7be911a4ebbb4e4d41963d4d260/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/user/repo/blob/93997d2ef089c7be911a4ebbb4e4d41963d4d260/tex/proofs/#1.ipynb}{\raisebox{-0.1em}{\prooficon}}}

### 2019-08-02 tex/ms.tex
more words

The observed spectrum is a complex function
of spatial, spectral, temporal, and velocity terms. The goal in this
section is to deconvolve each of these terms to make solving the integral
in Equation~(\ref{eq:intro:F}) easier.
%
The first thing we will do is to express Equation~(\ref{eq:intro:Ixi})
as a convolution:
%
\begin{align}
    \label{eq:deconv:convolution}
    I(\lnlam, \x, t, \Dargs) &= 
        I(\lnlam_0, \x, t) 
        * 
        \delta\big(\lnlam_0 + \D(\x, \Dargs)\big)
\end{align}
%
where $\delta$ is the
Dirac delta function and
$*$ denotes the linear convolution operator,
returning a function of the shifted log wavelength, 
$\lnlam = \lnlam_0 + \D$. The convolution with the delta function
has the effect of shifting the spectrum by an amount $\D$
onto the observed wavelength grid.

@@

\subsection{Computing $\g$}
\label{sec:g}
%
In the case that the star's rotation is rigid (i.e., differential rotation
is negligible) and other effects such as convective blueshift
may be ignored, the radial velocity at any point on the surface is 
simply proportional to the distance to the axis of rotation. Without loss
of generality, if we assume the axis of rotation lies along the $y-z$ plane,
we have
%
\begin{align}
    \beta(\x, \Dargs) = \frac{\omega \sin i \, x}{c}
\end{align}
%
and
%
\begin{align}
    \label{eq:g:D}
    \D(\x, \Dargs) &= 
        \frac{1}{2}\ln\left( 
            \dfrac{1 + \dfrac{\omega \sin i \, x}{c}}
                 {1 - \dfrac{\omega \sin i \, x}{c}}
        \right)
    \quad ,
\end{align}
%
where the parameters $\Dargs$ of the Doppler field are the
angular velocity, $\omega$, and the inclination of the star with
respect to $\hat{y}$, $i$.

@@

This completes our derivation of the analytic expression for the
observed spectrum (Equation~\ref{eq:deconv:F}).

@@

\section{Performance}
\label{sec:performance}
Evaluation of the model via Equation~(\ref{eq:deconv:F}) requires 
first the rotation of the $(\lmax + 1)^2$
$\g$ functions (each of length $W$) to the correct stellar 
rotational phase, an operation that scales approximately
as $W \lmax^3$,
followed by $(\lmax + 1)^2$
linear convolutions of the rotated $\g$ functions with the 
spectral components (each of length $K$), which scales
approximately as $W K \lmax^2$. Since typically $K \gg \lmax$,
the latter computation dominates and the entire evaluation 
takes $\mathcal{O}(W K \lmax^2)$ operations.

@@

\begin{figure}[p!]
    \begin{centering}
    \includegraphics[width=1.0\linewidth]{figures/linalg.pdf}
    \oscaption{linalg}{%
        The Doppler design matrix, constructed from a grid of Toeplitz
        matrices, each of shape $K \times (K + W - 1)$, where $K$
        is the number of wavelength bins in the observed spectrum and
        $W$ is the width of the convolution kernels.
        The $N$ columns of $\D$ correspond
        to the Toeplitz matrices for each of the $N$ spherical harmonics
        (shown at the top);
        the $M$ rows correspond to the Toeplitz matrices rotated to each
        of the $M$ stellar phases observed (indicated graphically 
        to the left of $\bvec{D}$).
        \label{fig:linalg}
    }
    \end{centering}
\end{figure}

>>>>

The observed spectrum is a complex function
of spatial, spectral, temporal, and velocity terms. The goal in this
section is to deconvolve each of these terms to make solving the integral
in Equation~(\ref{eq:intro:F}) easier.
%
The first thing we will do is to express Equation~(\ref{eq:intro:Ixi})
as a convolution:
%
\begin{align}
    \label{eq:deconv:convolution}
    I(\lnlam, \x, t, \Dargs) &= 
        I(\lnlam_0, \x, t) 
        * 
        \delta\big(\lnlam_0 + \D(\x, \Dargs)\big)
\end{align}
%
where $\delta$ is the
Dirac delta function and
$*$ denotes the linear convolution operator, defined for
two arbitrary functions $f$ and $h$ as the integral
%
\begin{align}
    \label{eq:deconv:convolution_def}
    (f * h)(t) \equiv \int_{-\infty}^\infty f(\tau) h(t - \tau) d\tau
\end{align}
%
for some independent coordinate $t$ and dummy parameter $\tau$.
%
The convolution of $I(\lnlam_0)$ with a delta function
has the effect of shifting the spectrum by an amount $\D$, returning
a function of the shifted (observed) log wavelength, 
$\lnlam = \lnlam_0 + \D$.

@@

\subsection{Computing $\g^\top$}
\label{sec:g}
%
In the case that the star's rotation is rigid (i.e., differential rotation
is negligible) and other effects such as convective blueshift
may be ignored, the radial velocity at any point on the surface is 
simply proportional to the distance to the axis of rotation. Without loss
of generality, if we assume the axis of rotation lies along the $y-z$ plane,
we have
%
\begin{align}
    \beta(\x, \Dargs) = \frac{\omega \sin i \, x}{c}
\end{align}
%
and
%
\begin{align}
    \label{eq:g:D}
    \D(\x, \Dargs) &= 
        \frac{1}{2}\ln\left( 
            \dfrac{1 + \dfrac{\omega \sin i \, x}{c}}
                 {1 - \dfrac{\omega \sin i \, x}{c}}
        \right)
    \quad ,
\end{align}
%
where the parameters $\Dargs$ of the Doppler field are the
angular velocity, $\omega$, and the inclination of the star with
respect to $\hat{y}$, $i$.

@@

This completes our derivation of the analytic expression for the
observed spectrum (Equation~\ref{eq:deconv:F}). As we show later,
this equation allows us to compute the forward model for the
spectrum very efficiently, particularly if we make use of Fast-Fourier
Transform (FFT) -based convolution techniques. However, a convolution
can also be represented as a matrix dot product, which we can use to
express the model for the observed spectrum as a purely linear
operation on the map decomposition $\alm(\lnlam_0)$. In the following section
we derive this expression and show how it can be used to obtain
fast, analytic solutions to the Doppler problem.

@@

\begin{figure}[ht!]
    \begin{centering}
    \includegraphics[width=0.8\linewidth]{figures/linalg.pdf}
    \oscaption{linalg}{%
        The Doppler design matrix $\bvec{D}$ (Equation~\ref{eq:linear:f}), 
        constructed from a grid of Toeplitz
        matrices, each of shape $K \times (K + W - 1)$, where $K$
        is the number of wavelength bins in the observed spectrum and
        $W$ is the width of the convolution kernels.
        The $N$ columns of $\bvec{D}$ correspond
        to the Toeplitz matrices for each of the $N$ spherical harmonics
        (shown at the top);
        the $M$ rows correspond to the Toeplitz matrices rotated to each
        of the $M$ stellar phases observed (indicated graphically 
        to the left of $\bvec{D}$).
        \label{fig:linalg}
    }
    \end{centering}
\end{figure}

Although the convolution operator (Equation~\ref{eq:deconv:convolution_def})
is defined for continuous functions, the problem of Doppler imaging deals with
discrete measurements of a spectrum in bins of log wavelength. Provided our
wavelength grid is fine enough, we may therefore approximate 
Equation~(\ref{eq:deconv:convolution_def}) with a discrete convolution.
For discrete arrays $\bvec{f}$ and $\bvec{h}$, we have
%
\begin{align}
    \label{eq:linear:convolution_def}
    \bvec{f} * \bvec{h} = \bvec{T}(\bvec{f}) \, \bvec{h}
\end{align}
%
where $\bvec{T}(\bvec{f})$ is a Toeplitz matrix, a matrix whose diagonals
are constant from top left to bottom right. In this case, the diagonals
are constructed from the values of $\bvec{f}$. If $\bvec{f}$ has length
$N$, element $f_n$ is placed everywhere along the $k^\mathrm{th}$ diagonal 
of $\bvec{T}$ for $k = -N / 2 + n$; all other entries of $\bvec{T}$ are 
set to zero.

Given this formulation, we may re-write Equation~(\ref{eq:deconv:F}) as a
purely linear operation on $\alm(\lnlam_0)$:
%
\begin{align}
    \label{eq:linear:f}
    \bvec{f}
    &=
    \bvec{D}(\g^\top, \R)
    \,
    \alm
    \quad,
\end{align}
%
where we introduce the \emph{Doppler design matrix} $\bvec{D}$, constructed
by concatenating the Toeplitz matrices for each of the components of the
convolution kernel $\g^\top(\lnlam, \Dargs)$ after rotating it (via $\R(t)$) 
to the correct stellar phase.

\section{Validation}
\label{sec:validation}

Words here. Figure~\ref{fig:spot} and Figure~\ref{fig:compare}.

@@

\section{Performance}
\label{sec:performance}

Evaluation of the model via Equation~(\ref{eq:deconv:F}) requires 
first the rotation of the $(\lmax + 1)^2$
$\g$ functions (each of length $W$) to the correct stellar 
rotational phase, an operation that scales approximately
as $W \lmax^3$,
followed by $(\lmax + 1)^2$
linear convolutions of the rotated $\g$ functions with the 
spectral components (each of length $K$), which scales
approximately as $W K \lmax^2$. Since typically $K \gg \lmax$,
the latter computation dominates and the entire evaluation 
takes $\mathcal{O}(W K \lmax^2)$ operations.
